---
title: CS5014 Machine Learning  
subtitle: Lecture 2 Maths background review 
author: "Lei Fang"
date: 28 Jan 2021
output: 
  beamer_presentation:
#    keep_tex: true
#    toc: true
#    slide_level: 3
    includes:
      in_header: sta-beamer-header-simple.txt
#      after_body: ~/Dropbox/teaching/table-of-contents.txt
---


```{r setup, include=FALSE}
library(rgl)
knitr::opts_chunk$set(webgl = hook_webgl,
                      echo = FALSE,
                      fig.align = 'center',
                      message =FALSE,
                      warning = FALSE
                      )
```

```{r, include = FALSE}
library(tidyverse)
library(reticulate)
# py_config()
```
# Introduction

## So why this review session ?
Maths is essential ...

- for CS5014 (all other (data) *science* modules)
- we use a lot of maths 
  - rigorous and concise way of communicating results 
  - help us understand why *and* why not algorithms work

\pause 
\bigskip Review some key concepts 

- not complete; just a small subset 
- set the level of math maturity we expect 

\pause
\bigskip Self-assessment for yourself
  
- find the weak or rusty area
- do self studies 

## Mathematics for machine learning 

Linear algebra

- leap forward from scalar maths e.g. $a + b = c$ 
  - $a,b,c$are scalars: $1+2 = 3$
  - $a,b,c$are vectors: $\begin{bmatrix}
           1 \\
           2 
         \end{bmatrix} + \begin{bmatrix}
           2 \\
           1 
         \end{bmatrix} = \begin{bmatrix}
           3 \\
           3 
         \end{bmatrix}$

\pause
\bigskip Probability theory and statistics 

- study of uncertainty: uncertainty is the norm
  - e.g. rain tomorrow? blood pressure measurement (reading error)? 
- how to generalise your results 
  - from one sample to the universe: vaccine trial

\pause 
\bigskip Calculus 

- study of continuous (real-valued) functions (using approximation *polynomial*)
  - $y= sin(x)$ is well approximated by $y=x$ when $x \approx 0$ 
- useful when we do optimisation 



# Linear Algebra
## Linear algebra: Basic concepts

- vectors
\bigskip
- norms and distances
\bigskip
- linear independence
\bigskip
- matrices, linear transformation
\bigskip
- matrix arithmetics
\bigskip
- rank, determinant, trace


## Vector 

A vector is a collection of $n$ salars

- $\vv{a} \in R^n$, default option is column vector i.e. $n\times 1$
- represents a **displacement** from $\vv{0}$ to $\vv{a}$ in $R^n$
- e.g. $\vv{a} = \begin{bmatrix}
           2 \\
           1 
         \end{bmatrix}, \textcolor{red}{\vv{b} = \begin{bmatrix}
           1 \\
           2 
         \end{bmatrix}}$ (or $\vv{a} = [2, 1]^T$ to save space)
\bigskip   
```{r, figures-side, fig.show="hold", out.width="49%"}
library(matlib)
#setting up the plot
xlim <- c(-1,4)
ylim <- c(-1,4)
par(mar=c(3,3,3,0.1)+.1)
plot(xlim, ylim, type="n", xlab="", ylab="", cex.lab=2.5, cex.axis=2.5, cex.main=2.5, cex.sub=2.5)
grid()
# define some vectors
a=c(2,1)
b=c(1,2)
# plot the vectors
vectors(b, labels="b", pos.lab=4, cex.lab = 3, col="red")
vectors(a, labels="a", pos.lab=4, cex.lab = 3)

# vec <- rbind(diag(3), c(1,1,1))
# rownames(vec) <- c("X", "Y", "Z", "J")
# open3d()
# vectors3d(vec, color=c(rep("black",3), "red"), lwd=2)
# # draw the XZ plane, whose equation is Y=0
# planes3d(0, 0, 1, 0, col="gray", alpha=0.2)
# vectors3d(c(1,1,0), col="green", lwd=2)
# plot(xlim, ylim, type="n", xlab="", ylab="", cex.lab=2.5, cex.axis=2.5, cex.main=3.5, cex.sub=2.5, main = "all the b vectors are the same")
# grid()
# vectors(b, labels="b", pos.lab=4, cex.lab = 3, col="red")
# displacement = c(2,0)
# vectors(b+displacement, labels="b", pos.lab=4, cex.lab = 3, origin = displacement, col="red")
# displacement = c(1,0)
# vectors(b+displacement, labels="b", pos.lab=4, cex.lab = 3, origin = displacement, col="red")
# displacement = c(2,-1)
# vectors(b+displacement, labels="b", pos.lab=4, cex.lab = 3, origin = displacement, col="red")
# displacement = c(0,1)
# vectors(b+displacement, labels="b", pos.lab=4, cex.lab = 3, origin = displacement, col="red")
# displacement = c(0,2)
# vectors(b+displacement, labels="b", pos.lab=4, cex.lab = 3, origin = displacement, col="red")
# vector a+b starting from a is equal to b.
#vectors(a+b, labels="b", pos.lab=4, frac.lab=.5, origin=a, col="green")
```

## experiment 

```{r, test-rgl, webgl=TRUE}
x <- sort(rnorm(1000))
y <- rnorm(1000)
z <- rnorm(1000) + atan2(x,y)
plot3d(x, y, z, col = rainbow(1000))
```

<!-- A vector is $$\vv{x} $$ -->
<!-- ## R Markdown -->

<!-- This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>. -->

<!-- When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. -->

<!-- ## Slide with Bullets -->

<!-- - Bullet 1 -->
<!-- - Bullet 2 -->
<!-- - Bullet 3 -->

<!-- ## Slide with R Output -->

<!-- ```{r cars, echo = TRUE} -->
<!-- summary(cars) -->
<!-- ``` -->

<!-- ## Slide with Plot -->

<!-- ```{r pressure} -->
<!-- plot(pressure) -->
<!-- ``` -->

